---
title: "UCSCXenaTools: Download Public Cancer Genomic Data from UCSC Xena Hubs"
author: "Shixiang Wang \\

        ShanghaiTech. University"
date: "`r Sys.Date()`"

output:
  prettydoc::html_pretty:
    toc: true
    theme: cayman
    highlight: github
  pdf_document:
    toc: true
vignette: >
  %\VignetteIndexEntry{download publical data in batch quickly and easily}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

**Current Version: 0.2.1**

`UCSCXenaTools` is a R package download and explore data from **UCSC Xena data hubs**, which are

> A collection of UCSC-hosted public databases such as TCGA, ICGC, TARGET, GTEx, CCLE, and others. Databases are normalized so they can be combined, linked, filtered, explored and downloaded.
>
> -- [UCSC Xena](https://xena.ucsc.edu/)


## Installation

You can install UCSCXenaTools from github with:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("ShixiangWang/UCSCXenaTools", build_vignettes = TRUE)
```

Read this vignettes.

```{r, eval=FALSE}
browseVignettes("UCSCXenaTools")
```


## Data Hub List

All datasets are available at <https://xenabrowser.net/datapages/>.

Currently, `UCSCXenaTools` support all 7 data hubs of UCSC Xena.

* [UCSC Public Hub](https://xenabrowser.net/datapages/?host=https%3A%2F%2Fucscpublic.xenahubs.net): <https://ucscpublic.xenahubs.net>
* [TCGA Hub](https://xenabrowser.net/datapages/?host=https%3A%2F%2Ftcga.xenahubs.net): <https://tcga.xenahubs.net>
* [GDC Xena Hub](https://xenabrowser.net/datapages/?host=https%3A%2F%2Fgdc.xenahubs.net): <https://gdc.xenahubs.net>
* [ICGC Xena Hub](https://xenabrowser.net/datapages/?host=https%3A%2F%2Ficgc.xenahubs.net): <https://icgc.xenahubs.net>
* [Pan-Cancer Atlas Hub](https://xenabrowser.net/datapages/?host=https%3A%2F%2Fpancanatlas.xenahubs.net): <https://pancanatlas.xenahubs.net>
* [GA4GH (TOIL) Hub](https://xenabrowser.net/datapages/?host=https%3A%2F%2Ftoil.xenahubs.net): <https://toil.xenahubs.net>
* [Treehouse Hub](https://xenabrowser.net/datapages/?host=https%3A%2F%2Fxena.treehouse.gi.ucsc.edu): <https://xena.treehouse.gi.ucsc.edu>

If the `API` changed, please remind me by email to <w_shixiang@163.com> or open an issue on [GitHub](https://github.com/ShixiangWang/UCSCXenaTools/issues).

## Usage

Download UCSC Xena Datasets and load them into R by `UCSCXenaTools` is a workflow in `generate`, `filter`, `query`, `download` and `prepare` 5 steps, which are implemented as `XenaGenerate`, `XenaFilter`, `XenaQuery`, `XenaDownload` and `XenaPrepare`, respectively. They are very clear and easy to use and combine with other packages like `dplyr`.

The following use clinical data download of LUNG, LUAD, LUSC from TCGA (hg19 version) as an example.


### XenaData data.frame

Begin from version `0.2.0`, `UCSCXenaTools` use a `data.frame` object (built in package) `XenaData` to generate an instance of `XenaHub` class, which communicate with API of UCSC Xena Data Hubs.

You can load `XenaData` after loading `UCSCXenaTools` into R.

```{r}
library(UCSCXenaTools)
data(XenaData)

head(XenaData)
```



### Generate a XenaHub object

This can be implemented by `XenaGenerate` function, which generate `XenaHub` object from `XenaData` data frame.

```{r}
XenaGenerate()
```

We can set `subset` argument to narrow datasets.

```{r}
XenaGenerate(subset = XenaHostNames=="TCGA")
```

>You can use `XenaHub()` to generate a `XenaHub` object for API communication, but it is not recommended. 

It's possible to explore `hosts()`, `cohorts()` and `datasets()`.

```{r}
xe = XenaGenerate(subset = XenaHostNames=="TCGA")
# get hosts
hosts(xe)
# get cohorts
head(cohorts(xe))
# get datasets
head(datasets(xe))
```

Pipe operator `%>%` can also be used here.


```
> library(tidyverse)
> XenaData %>% filter(XenaHostNames == "TCGA", grepl("BRCA", XenaCohorts), grepl("Path", XenaDatasets)) %>% XenaGenerate()
class: XenaHub 
hosts():
  https://tcga.xenahubs.net
cohorts() (1 total):
  TCGA Breast Cancer (BRCA)
datasets() (4 total):
  TCGA.BRCA.sampleMap/Pathway_Paradigm_mRNA_And_Copy_Number
  TCGA.BRCA.sampleMap/Pathway_Paradigm_RNASeq
  TCGA.BRCA.sampleMap/Pathway_Paradigm_RNASeq_And_Copy_Number
  TCGA.BRCA.sampleMap/Pathway_Paradigm_mRNA
```

### Filter 

There are too many datasets, we filter them by `XenaFilter` function.

Regular expression can be used to filter XenaHub object to what we want.


```{r}
(XenaFilter(xe, filterDatasets = "clinical") -> xe2)
```


Then select `LUAD`, `LUSC` and `LUNG` 3 datasets.

```{r}
XenaFilter(xe2, filterDatasets = "LUAD|LUSC|LUNG") -> xe2
```

Pipe can be used here.

```
suppressMessages(require(dplyr))

xe %>% 
    filterXena(filterDatasets = "clinical") %>% 
    filterXena(filterDatasets = "luad|lusc|lung")
## class: XenaHub 
## hosts():
##   https://tcga.xenahubs.net
## cohorts() (39 total):
##   (unassigned)
##   TCGA Acute Myeloid Leukemia (LAML)
##   TCGA Adrenocortical Cancer (ACC)
##   ...
##   TCGA Thyroid Cancer (THCA)
##   TCGA Uterine Carcinosarcoma (UCS)
## datasets() (3 total):
##   TCGA.LUSC.sampleMap/LUSC_clinicalMatrix
##   TCGA.LUNG.sampleMap/LUNG_clinicalMatrix
##   TCGA.LUAD.sampleMap/LUAD_clinicalMatrix
```

### Query

Create a query before download data

```{r}
xe2_query = XenaQuery(xe2)
xe2_query
```

### Download

Default, data will be downloaded to `XenaData` directory under system temp directory. You can specify the path.

If the data exists, command will not run to download them, but you can force it by `force` option.

```{r}
xe2_download = XenaDownload(xe2_query)
## not run
#xe2_download = XenaDownload(xe2_query, destdir = "E:/Github/XenaData/test/")
```

> Note fileNames transfromed from datasets name and / chracter all changed to __ character.

### Prepare

There are 4 ways to prepare data to R.

```
# way1:  directory
cli1 = XenaPrepare("E:/Github/XenaData/test/")
names(cli1)
## [1] "TCGA.LUAD.sampleMap__LUAD_clinicalMatrix.gz"
## [2] "TCGA.LUNG.sampleMap__LUNG_clinicalMatrix.gz"
## [3] "TCGA.LUSC.sampleMap__LUSC_clinicalMatrix.gz"
```

```
# way2: local files
cli2 = XenaPrepare("E:/Github/XenaData/test/TCGA.LUAD.sampleMap__LUAD_clinicalMatrix.gz")
class(cli2)
## [1] "tbl_df"     "tbl"        "data.frame"

cli2 = XenaPrepare(c("E:/Github/XenaData/test/TCGA.LUAD.sampleMap__LUAD_clinicalMatrix.gz",
                     "E:/Github/XenaData/test/TCGA.LUNG.sampleMap__LUNG_clinicalMatrix.gz"))
class(cli2)
## [1] "list"
names(cli2)
## [1] "TCGA.LUAD.sampleMap__LUAD_clinicalMatrix.gz"
## [2] "TCGA.LUNG.sampleMap__LUNG_clinicalMatrix.gz"
```

```
# way3: urls
cli3 = XenaPrepare(xe2_download$url[1:2])
names(cli3)
## [1] "LUSC_clinicalMatrix.gz" "LUNG_clinicalMatrix.gz"
```

```{r}
# way4: xenadownload object
cli4 = XenaPrepare(xe2_download)
names(cli4)
```

### SessionInfo

```{r}
sessionInfo()
```

## Acknowledgement

This package is based on [XenaR](https://github.com/mtmorgan/XenaR), thanks [Martin Morgan](https://github.com/mtmorgan) for his work.

## LICENSE

GPL-3

please note, code from XenaR package under Apache 2.0 license.

## ToDo

* Shinny
* More easier download workflow
