---
title: "UCSCXenaTools: an R package for Accessing Genomics Data from UCSC Xena platform, from Cancer Multi-omics to Single-cell RNA-seq"
author: "Shixiang Wang \\

        ShanghaiTech. University"
date: "`r Sys.Date()`"

output:
  prettydoc::html_pretty:
    toc: true
    theme: cayman
    highlight: github
  pdf_document:
    toc: true
vignette: >
  %\VignetteIndexEntry{Basic usage}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


**UCSCXenaTools** is an R package for accessing genomics data from UCSC Xena platform, 
from cancer multi-omics to single-cell RNA-seq. 
Public omics data from UCSC Xena are supported through [**multiple turn-key Xena Hubs**](https://xenabrowser.net/datapages/), which are a collection of UCSC-hosted public databases such as TCGA, ICGC, TARGET, GTEx, CCLE, and others. Databases are normalized so they can be combined, linked, filtered, explored and downloaded.

**Who is the target audience and what are scientific applications of this package?**

* Target Audience: cancer and clinical researchers, bioinformaticians
* Applications: genomic and clinical analyses

## Installation

Install stable release from CRAN with:

```{r, eval=FALSE}
install.packages("UCSCXenaTools")
```

You can also install devel version of **UCSCXenaTools** from github with:

```{r gh-installation, eval = FALSE}
# install.packages("remotes")
remotes::install_github("ShixiangWang/UCSCXenaTools")
```

If you want to build vignette in local, please add two options:

```{r, eval=FALSE}
remotes::install_github("ShixiangWang/UCSCXenaTools", build_vignettes = TRUE, dependencies = TRUE)
```

The minimum versions to run the vignette is `1.2.4`. 
[GitHub Issue](https://github.com/ShixiangWang/UCSCXenaTools/issues) is a place for discussing any problem.

## Data Hub List

All datasets are available at <https://xenabrowser.net/datapages/>.

Currently, **UCSCXenaTools** supports 10 data hubs of UCSC Xena.

* UCSC Public Hub: <https://ucscpublic.xenahubs.net>
* TCGA Hub: <https://tcga.xenahubs.net>
* GDC Xena Hub: <https://gdc.xenahubs.net>
* ICGC Xena Hub: <https://icgc.xenahubs.net>
* Pan-Cancer Atlas Hub: <https://pancanatlas.xenahubs.net>
* GA4GH (TOIL) Hub: <https://toil.xenahubs.net>
* Treehouse Hub: <https://xena.treehouse.gi.ucsc.edu>
* PCAWG Hub: <https://pcawg.xenahubs.net>
* ATAC-seq Hub: <https://atacseq.xenahubs.net>
* Singel Cell Xena hub: <https://singlecell.xenahubs.net>

If any url of data hub is changed or a new data hub is online, please remind me by emailing to <w_shixiang@163.com> or [opening an issue on GitHub](https://github.com/ShixiangWang/UCSCXenaTools/issues).


## Usage

Download UCSC Xena datasets and load them into R by **UCSCXenaTools** is a workflow with `generate`, `filter`, `query`, `download` and `prepare` 5 steps, which are implemented as `XenaGenerate`, `XenaFilter`, `XenaQuery`, `XenaDownload` and `XenaPrepare` functions, respectively. They are very clear and easy to use and combine with other packages like `dplyr`.

To show the basic usage of **UCSCXenaTools**, we will download clinical data of LUNG, LUAD, LUSC from TCGA (hg19 version) data hub.

### XenaData data.frame

**UCSCXenaTools** uses a `data.frame` object (built in package) `XenaData` to generate an instance of `XenaHub` class, which records information of all datasets of UCSC Xena Data Hubs.

You can load `XenaData` after loading `UCSCXenaTools` into R.

```{r}
library(UCSCXenaTools)
data(XenaData)

head(XenaData)
```

### Workflow

Select datasets.

```{r}
# The options in XenaFilter function support Regular Expression
XenaGenerate(subset = XenaHostNames=="tcgaHub") %>% 
  XenaFilter(filterDatasets = "clinical") %>% 
  XenaFilter(filterDatasets = "LUAD|LUSC|LUNG") -> df_todo

df_todo
```

Sometimes we only know some keywords, `XenaScan()` can be used to scan all rows to detect if 
the keywords exist in `XenaData`.


```{r}
x1 = XenaScan(pattern = 'Blood')
x2 = XenaScan(pattern = 'LUNG', ignore.case = FALSE)

x1 %>%
    XenaGenerate()
x2 %>%
    XenaGenerate()
```


Query and download.

```{r}
XenaQuery(df_todo) %>%
  XenaDownload() -> xe_download
```

Prepare data into R for analysis.

```{r}
cli = XenaPrepare(xe_download)
class(cli)
names(cli)
```


### Browse datasets

Create two XenaHub objects:

* `to_browse` - a XenaHub object containing a cohort and a dataset.
* `to_browse2` - a XenaHub object containing 2 cohorts and 2 datasets.

```{r}
XenaGenerate(subset = XenaHostNames=="tcgaHub") %>%
    XenaFilter(filterDatasets = "clinical") %>%
    XenaFilter(filterDatasets = "LUAD") -> to_browse

to_browse

XenaGenerate(subset = XenaHostNames=="tcgaHub") %>%
    XenaFilter(filterDatasets = "clinical") %>%
    XenaFilter(filterDatasets = "LUAD|LUSC") -> to_browse2

to_browse2
```

`XenaBrowse()` function can be used to browse dataset/cohort links using your default web browser.
At default, this function limits one dataset/cohort for preventing user to open too many links at once. 

```{r,eval=FALSE}
# This will open you web browser
XenaBrowse(to_browse)

XenaBrowse(to_browse, type = "cohort")
```

```{r, error=TRUE}
# This will throw error
XenaBrowse(to_browse2)

XenaBrowse(to_browse2, type = "cohort")
```

When you make sure you want to open multiple links, you can set `multiple` option to `TRUE`.

```{r, eval=FALSE}
XenaBrowse(to_browse2, multiple = TRUE)
XenaBrowse(to_browse2, type = "cohort", multiple = TRUE)
```

## More usages

The core functionality has been described above. 
I write more usages about this package in my website but not here
because sometimes package check will fail due to internet problem.

- [Introduction and basic usage of UCSCXenaTools](https://shixiangwang.github.io/home/en/tools/ucscxenatools-intro/) - [PDF](https://shixiangwang.github.io/home/en/tools/ucscxenatools-intro.pdf)
- [APIs of UCSCXenaTools](https://shixiangwang.github.io/home/en/tools/ucscxenatools-api/) - [PDF](https://shixiangwang.github.io/home/en/tools/ucscxenatools-api.pdf)

## Citation

Cite me by the following paper.

```
Wang, Shixiang, et al. "The predictive power of tumor mutational burden 
    in lung cancer immunotherapy response is influenced by patients' sex." 
    International journal of cancer (2019).
```

Cite UCSC Xena by the following paper. 

```
Goldman, Mary, et al. "The UCSC Xena Platform for cancer genomics data 
    visualization and interpretation." BioRxiv (2019): 326470.
```

## Acknowledgments

This package is based on [XenaR](https://github.com/mtmorgan/XenaR), thanks [Martin Morgan](https://github.com/mtmorgan) for his work.
