---
title: "UCSCXenaTools: R API for UCSC Xena Hubs"
author: "Shixiang Wang \\

        ShanghaiTech. University"
date: "`r Sys.Date()`"

output:
  prettydoc::html_pretty:
    toc: true
    theme: cayman
    highlight: github
  pdf_document:
    toc: true
vignette: >
  %\VignetteIndexEntry{R API for UCSC Xena Hubs}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette gives users the summary information of API functions provided by **UCSCXenaTools** for UCSC Xena. 

Before using API, user should know some concepts about Xena elements. Following description is copied from [xenaPython `__init__.py`](https://github.com/ucscXena/xenaPython/blob/master/xenaPython/__init__.py).

> Data rows are associated with "sample" IDs.
>
> Sample IDs are unique within a "**cohort**".
>s
> A "**dataset**" is a particular assay of a cohort, e.g. gene expression.
>
> Datasets have associated metadata, specifying their data type and cohort.
>
> There are three primary data types: **dense matrix** (samples by probes), **sparse** (sample, position, variant), and **segmented** (sample, position, value).
> 
> Dense matrices can be genotypic or phenotypic. Phenotypic matrices have associated **field metadata** (descriptive names, codes, etc.).
Genotypic matricies may have an associated **probeMap**, which maps probes to genomic locations. If a matrix has hugo probeMap, the probes themselves are gene names. Otherwise, a probeMap is used to map a gene location to a set of probes.

## API categories

API functions can be divided into two classes: **lower API functions** and **higher API functions**. 
They have following difference:

- The main difference between them is that the target of higher API functions is `XenaHub` object, which is
a S4 class built in R. While the targets of lower API functions are Xena hub urls, cohort names or 
dataset names with character format. The `XenaHub` object can provide more uniform operation methods and can be
used to download corresponding datasets quickly and easily 
(detail see [another vignette](https://shixiangwang.github.io/UCSCXenaTools/)).
- Lower API functions are not registered in package 
[NAMESPACE](https://github.com/ShixiangWang/UCSCXenaTools/blob/master/NAMESPACE), so user may not access them
after `library(UCSCXenaTools)`, user need to use `UCSCXenaTools:::fun_name` instead.
- Lower API functions have no help pages, so user cannot find any description about them in R, which means
you cannot use `?fun_name` to get help. However, API report part in this vignette shows all avaiable API functions and 
their short description.
- Higher API functions are built on lower API functions, they return more meaningful and easy results for operation. Most of lower API functions return nested lists as results, user need to tidy them before using them in next step.


### Lower API functions

Lower API functions also have 2 classes: 

- one is generated from 
[`.xq` files](https://github.com/ShixiangWang/UCSCXenaTools/tree/master/inst/queries), 
function names all start with `.p_`. All `.xq` files are copied from [xenaPython](https://github.com/ucscXena/xenaPython) package, which is official Python API for Xena.
These functions are dynamicly created when **UCSCXenaTools** loaded. Their names are given as following: 

    ```{r include=FALSE}
    #devtools::load_all("~/Repo/UCSCXenaTools/")
    library(UCSCXenaTools)
    all_vs = ls(envir = as.environment("package:UCSCXenaTools"),
                all.names = TRUE)
    ```
    
    ```{r echo=FALSE}
    grep("^\\.p", all_vs, value = TRUE)
    ```

- the other one is created in package. The function names all start with `.`, are given as following:

    ```{r, echo=F}
    funs_low_in = c(
        '.host_cohorts',
        '.cohort_datasets',
        '.cohort_datasets_count',
        '.cohort_samples_each',
        '.cohort_samples_any',
        '.cohort_samples_all',
        '.dataset_samples_each',
        '.dataset_samples_any',
        '.dataset_samples_all'
    )
    funs_low_in
    ```


I don't know how to write these query sentence for Xena Hubs. So here I want to
say thanks to authors of [**xenaPython**](https://github.com/ucscXena/xenaPython) 
and [**xenaR**](https://github.com/mtmorgan/XenaR) packages.

## API report


```{r echo=FALSE, message=FALSE, warning=FALSE}
#devtools::load_all("~/Repo/UCSCXenaTools/")
library(dplyr)
library(UCSCXenaTools)
load(system.file("api.RData", package = "UCSCXenaTools"))
api_df = arrange(api_df, Level)
knitr::kable(api_df, caption = "API functions in UCSCXenaTools")
```

Of note, I don't know test all functions generated from `.xq` files, most of them works.
Sometimes functions return you errors or `list()` may caused by invaild format or bad network, 
you should try more times. If you make sure there are problems/errors in query procedure,
you can check corresponding query variables:

```{r echo=FALSE}
    grep("^\\.xq", all_vs, value = TRUE)
```

For example, you'd like to check `.p_all_cohorts` function, you can take a look
at `.xq_all_cohorts` object.

```{r}
.xq_all_cohorts
```

`cat` it may give you more easy-to-read format.

```{r}
cat(.xq_all_cohorts)
```

## Use case

More to do, if you have any suggestion, you can open issue on GitHub.


## LICENSE

GPL-3

Please note, code from XenaR package under Apache 2.0 license.
